% ============================================================
%  NHB_main.tex  —  versão compatível com o template Nature/NHB
% ============================================================

\documentclass{nature}             % classe Nature
\usepackage{graphicx}
\usepackage{amsmath, amssymb}
\usepackage{hyperref}
% Pacotes opcionais
\usepackage{authblk}               % se desejar afiliações numeradas
\usepackage{lineno}                % linha numerada em submissão
% \linenumbers                     % descomente para submissão

% ---------- Título, Autores, Afiliações ----------
\title{Symbolic Manifolds and Entropic Dynamics: \\ A Cognitive Topology of Mental States}

\author{Demetrios Chiuratto Agourakis$^{1,2}$}

\begin{document}

\maketitle

\begin{affiliations}
 \item Faculdade São Leopoldo Mandic, Campinas, SP, Brazil.
 \item Postgraduate Programme in Biomaterials and Regenerative Medicine, Pontifícia Universidade Católica de São Paulo (PUC‑SP), Brazil.
\end{affiliations}

% ---------- Abstract (≤150 palavras) ----------
\begin{abstract}
We introduce a symbolic-topological framework for cognition in which mental states evolve as trajectories $\gamma(t) = (\alpha, \kappa, E_r)$ on a three-dimensional manifold. Simulations reproduce neurotypical, gifted, twice-exceptional (2e), and collapse-prone regimes, revealing bifurcations and collapse–recovery dynamics. The model unifies entropy-driven brain theories with symbolic cognition and offers testable biomarkers, laying the groundwork for a *topological symbolic psychiatry*.
\end{abstract}
% ---------- Seções principais ----------
\section*{Introduction}
\input{sections/01_intro.tex}

\section*{Results}
\subsection*{A symbolic cognitive topology}
\input{sections/02_model_structure.tex}

\subsection*{Entropic dynamics in simulated manifolds}
\input{sections/03_simulation_results.tex}

\section*{Discussion}
\input{sections/04_discussion.tex}

% ---------- Methods ----------
\section*{Methods}
\input{sections/05_methods.tex}

% ---------- Data & Code availability ----------
\section*{Data availability}
All data (simulation scripts and generated output) are freely available in the public GitHub repository \url{https://github.com/agourakis82/entropic-symbolic-society} and archived under Zenodo DOI \href{https://doi.org/10.5281/zenodo.16682785}{10.5281/zenodo.16682785}.

\section*{Code availability}
Custom Python code used in the present study is provided alongside the data repository.

% ---------- AI assistance ----------
\section*{AI assistance}
This manuscript benefitted from \textbf{OpenAI GPT‑4o / o3‑pro} assistance exclusively for
(1) language polishing,
(2) structural organisation of sections, and
(3) LaTeX syntax verification.
No content was generated without subsequent critical review and approval by the author.
The author takes full responsibility for the integrity and accuracy of the work.

% ---------- Acknowledgements ----------
\section*{Acknowledgements}
The author thanks the Biomaterials and Regenerative Medicine graduate community at PUC‑SP for valuable feedback, and acknowledges Overleaf Premium and Zotero Better BibTeX for infrastructure support.

% ---------- Author contributions ----------
\section*{Author contributions}
D.C.A. conceived the study, developed the theoretical framework, performed simulations, analysed the results and wrote the manuscript.

% ---------- Competing interests ----------
\section*{Competing interests}
The author declares no competing interests.

% ---------- References ----------
\bibliographystyle{naturemag}      % estilo Nature
\bibliography{references}

% ---------- Extended Data / Supplementary ----------
% Use "extended data" figures/tables if NHB requests.
% \begin{extendeddatafigure}
% ...
% \end{extendeddatafigure}

\end{document}
